---
title: "Working with phyloseq and TreeSummarizedExperiment versions of the Global Patterns Dataset"
output: BiocStyle::html_document
vignette: >
  %\VignetteIndexEntry{Working with phyloseq and TreeSummarizedExperiment versions of the Global Patterns Dataset}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r opts, include = FALSE}
knitr::opts_chunk$set(
    collapse = TRUE,
    comment = "#>",
    message = FALSE,
    warning = FALSE
)
```

This vignette demonstrates how to create phylobar plots using either phyloseq or
TreeSummarizedExperiment objects. We'll work with the same dataset
(GlobalPatterns) stored in both formats.

## Setup

```{r setup}
library(phylobar)
library(phyloseq)
library(miaViz)
library(scater)
```

## phyloseq Inputs

We will first study how to use phylobar on phyloseq objects. The block below
sets the global pattern status and subsets to the Chlamydia subtree. We can
extract the relevant sample composition using the otu_table accessor function.

```{r phyloseq-phylo}
data(GlobalPatterns, package = "phyloseq")
chlamydiae <- subset_taxa(GlobalPatterns, Phylum == "Chlamydiae")
x_phylo <- t(otu_table(chlamydiae))
```

### Phylogenetic Tree Tree

Our first approach will use the phylogenetic tree to support hierarchical
interaction. The tree can be extracted from the phyloseq object using
`phy_tree`. We first subsetted to only the chlamydia subtree. There are now some
samples without any observed counts.  Therefore, we will remove the samples from
the view.

```{r handle-zeros}
tree_phylo <- phy_tree(chlamydiae)
x_phylo <- x_phylo[rowSums(x_phylo) > 0, ]
x_phylo <- x_phylo / rowSums(x_phylo)
```

We can now create our phylobar plot.

```{r phylobar-phyloseq}
phylobar(x_phylo, tree_phylo, width = 900)
```

### Using Taxonomic Hierarchy

This creates the same type of visualization but using a taxonomic hierarchy
rather than a phylogeny. We need some problem-specific preprocessing to deal
with the fact that some ancestor and descendant names look identical in this
dataset. The `add_prefix` helper ensures that the names are distinct between
parents and children by adding in the taxonomic rank to each name.

```{r phyloseq-taxonomy}
taxa <- tax_table(chlamydiae) |>
    as.data.frame()
taxa$ASV <- rownames(taxa)
taxa <- phylobar::add_prefix(taxa)
```

Now we can create a phylo tree object associated with the original taxonomy.
Note that this is not a binary tree, since several taxonomic categories
descended from a single node.

```{r construct_taxonomy}
tax_tree <- taxonomy_to_tree(taxa)
plot(tax_tree)
```

Given our new taxonomy and the earlier sample information, we can now create a
phylobar plot.

```{r phylobar-taxon}
phylobar(x_phylo, tax_tree, width = 900)
```

## TreeSummarizedExperiment Inputs

This example comes from the
[miaViz documentation](https://microbiome.github.io/miaViz/articles/miaViz.html)
and shows how to create a stacked bar tree from a TreeSummarizedExperiment.
We'll filter down to only those species that are present in at least one percent
of all samples. In this example we'll use a relative abundance transformation,
so that we left with a composition of our plot rather than a more general
stacked bar plot.

```{r tse-data-prep}
data(GlobalPatterns, package = "mia")
prev_species <- getPrevalent(GlobalPatterns, rank = "Species", detection = 0.01)
GlobalPatterns_tse <- GlobalPatterns[
    rowData(GlobalPatterns)$Species %in% prev_species,
]
GlobalPatterns_tse <- transformAssay(
    GlobalPatterns_tse,
    method = "relabundance"
)
```

Let's keep only those taxes that are observed in at least one sample. We also
need to filter the tree to reflect this reduced subset of taxa.

```{r tse-phylobar-x}
x_tse <- t(assay(GlobalPatterns_tse, "relabundance"))
x_tse <- x_tse[, colSums(x_tse) > 0]
tree_tse <- rowTree(GlobalPatterns_tse) |>
    keep.tip(colnames(x_tse))
```

With these inputs, we can generate the phylobar visualization. This view didn't
subset to the chlamydia class, which is why we have a larger tree here. But
hopefully it's clear that TreeSummarizedExperiment and phyloseq can essentially
be used interchangeably when constructing the necessary inputs for these
visualizations.

```{r phylobar-tse}
phylobar(x_tse, tree_tse, width = 900)
```

```{r session-info}
sessionInfo()
```